---
title: "HW3"
author: "Zakari L. Billo"
date: "2026-04-02"
output: pdf_document
---
Load Packages 

```{r, echo = T, message = FALSE, results='hide', warning=FALSE}
library(caret)
library(tidymodels)
library(discrim) 
library(MASS)
library(mlbench)
library(pROC)
library(klaR)
library(plotmo)
library(splines)
library(mgcv)
library(pdp)
library(earth)
library(tidyverse)
library(ggplot2)
library(ISLR)
library(glmnet)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 7,
  fig.align = "center",
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

Split the dataset into two parts: training data (70%) and test data (30%).

```{r}
dat <- read.csv("data/auto.csv",  na = c("NA", ".", "")) |>
  janitor::clean_names() |> na.omit() |>
  mutate(
    mpg_cat = as.factor(mpg_cat),
    origin = factor(origin, 
                    levels = c(1, 2, 3), 
                    labels = c("American", "European", "Japanese")),
  )

dat <- dat[dat$cylinders != 3, ]

summary(dat)

set.seed(1)
data_split <- initial_split(dat, prop = 0.7)

# Extract the training and test data
training_data <- training(data_split)
testing_data <- testing(data_split)

# set up cross validation control
ctrl <- trainControl(method = "cv", 10,
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE)
```

(a) Perform logistic regression analysis. Are there redundant predictors in your model? If so, identify them. If there are none, please provide an explanation.

```{r}
set.seed(1)

lr.fit <- train(
                mpg_cat ~ ., 
                data=training_data, 
                method="glm", 
                family="binomial",
                metric = "ROC",
                preProcess=c("center", "scale"), 
                trControl=ctrl)

summary(lr.fit)
print(lr.fit)
```

The final, unpenalized logistic regression model was $$\begin{aligned}
\text{logit}(p) &= 0.9181 + 0.9284 \cdot \text{cylinders}^* - 1.4203 \cdot \text{displacement}^* \\
&\quad + 2.2279 \cdot \text{horsepower}^* + 4.5912 \cdot \text{weight}^* + 0.2248 \cdot \text{acceleration}^* \\
&\quad - 1.7152 \cdot \text{year}^* - 0.4152 \cdot I(\text{origin} = \text{European}) \\
&\quad - 0.5572 \cdot I(\text{origin} = \text{Japanese})
\end{aligned}$$

Next, I used LASSO to examine the redundant features.

```{r}
set.seed(1)
lasso.fit <- train(
                  mpg_cat ~ .,
                  training_data,
                  method = "glmnet",
                  preProcess = c("center", "scale"),
                  tuneGrid = expand.grid(alpha = 1,
                                         lambda = exp(seq(10, -5, length = 100))),
                  trControl = ctrl) 

# Extract coefficients at the optimal lambda
lasso_coefs <- coef(lasso.fit$finalModel, s = lasso.fit$bestTune$lambda)
print(lasso_coefs)
```

The redundant predictors (shrunk to 0 by the L1 norm of LASSO) appear to be `acceleration` and `displacement`. An explanation for this is these measures of engine performance add dimensionality but do not each describe unique properties of the car. For example, `acceleration` is the derivative of velocity (the variable `displacement` over time, in a direction), but both do not inform the mileage that a car may have.  

The final penalized logistic regression model for LASSO was $$\begin{aligned}
\text{logit}(p) &= 0.6764 + 0.3225 \cdot \text{cylinders}^* + 0.8634 \cdot \text{horsepower}^* \\
&\quad + 3.0323 \cdot \text{weight}^* - 1.1614 \cdot \text{year}^* \\
&\quad - 0.0089 \cdot I(\text{origin} = \text{European}) - 0.1596 \cdot I(\text{origin} = \text{Japanese})
\end{aligned}$$

(b) Train a multivariate adaptive regression spline (MARS) model. Does the MARS
model improve prediction performance compared to logistic regression?

```{r, warning = FALSE}
set.seed(1)
mars_grid <- expand.grid(degree = 1:3, nprune = 2:25)

mars.fit <- train(mpg_cat ~ ., 
                   data = training_data,
                    method = "earth",
                    tuneGrid = mars_grid,
                    metric = "ROC",
                    trControl = ctrl)

plot(mars.fit)
mars.fit$bestTune
coef(mars.fit$finalModel)
print(mars.fit$finalModel)
summary(mars.fit$finalModel)
```

The final mars model selected was $$\begin{aligned}
\text{logit}(p) &= 8.8028 - 2.0454 \cdot h(6 - \text{cylinders}) + 1.1538 \cdot h(\text{displacement} - 119) \\
&\quad - 1.3189 \cdot h(\text{displacement} - 122) + 0.1743 \cdot h(\text{displacement} - 156) \\
&\quad - 0.0191 \cdot h(\text{displacement} - 232) - 0.0054 \cdot h(3525 - \text{weight}) \\
&\quad - 0.5812 \cdot h(\text{year} - 72)
\end{aligned}$$

```{r}
set.seed(1)
res <- resamples(list(Logistic = lr.fit,
                      Lasso = lasso.fit,
                      MARS = mars.fit))

summary(res)
bwplot(res, metric = "ROC", main = "CV ROC: Logistic Regression vs. LASSO vs. MARS")
```

Overall, it appears that the MARS model had a higher cross-validated ROC (0.984 vs. LASSO's 0.977 and LR's 0.974), Sensitivity (0.964 vs. LASSO's 0.936 and LR's 0.921) and Specificity (0.916 vs LASSO's 0.886 and LR's 0.901). Thus, the final cross-validated MARS model is an improvement over both the LASSO and Logistic regression model.

(c) Perform linear discriminant analysis using the training data. Plot the linear discriminant(s).

```{r, warning = FALSE}
# train LDA model with CV
# set up cross validation control
ctrl1 <- trainControl(method = "repeatedcv", repeats = 10,
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE)

set.seed(1)
lda.fit <- train(mpg_cat ~ ., 
                   data = training_data,
                   method = "lda",
                   metric = "ROC",
                   trControl = ctrl)

lda.fit$finalModel

# Fit a standalone LDA model just for the plot
lda.plot <- lda(mpg_cat ~ ., data = training_data)
plot(lda.plot)
```

The final LDA model was $$\begin{aligned}
\text{LD1} &= 0.5163 \cdot \text{cylinders} - 0.0015 \cdot \text{displacement} - 0.0069 \cdot \text{horsepower} \\
&\quad + 0.0012 \cdot \text{weight} + 0.0213 \cdot \text{acceleration} - 0.1191 \cdot \text{year} \\
&\quad - 0.1545 \cdot I(\text{origin} = \text{European}) - 0.4373 \cdot I(\text{origin} = \text{Japanese})
\end{aligned}$$

(d) Which model will you choose to predict the response variable? Plot its ROC curve
and report the AUC.

```{r}
set.seed(1)
res <- resamples(list(Logistic = lr.fit,
                      LASSO = lasso.fit,
                      MARS = mars.fit,
                      LDA = lda.fit))

summary(res)
bwplot(res, metric = "ROC", main = "CV ROC: Logistic vs. LASSO vs. MARS vs. LDA")
```

Based on 10-fold cross-validation, MARS is still the winning model.

```{r}
set.seed(1)
mars.probs <- predict(mars.fit,
                      newdata = testing_data,
                      type = "prob")[,2]

roc.mars <- roc(testing_data$mpg_cat, mars.probs)

auc(roc.mars)
plot(roc.mars, legacy.axes = TRUE,
     main = "ROC Curve: MARS Model")
```

The AUC for MARS is 0.9816.

(e) Select a probability threshold to classify observations and compute the confusion matrix for the final model you choose. Briefly interpret what the confusion matrix indicates about your model’s performance. (Note: We are looking for concise inter-pretations based on your own understanding. Do not use AI tools to answer this question.)

```{r}
test.probs <- predict(mars.fit, newdata = testing_data, type = "prob")
test.pred.prob <- test.probs[, "high"]

test.pred <- rep("low", length(test.pred.prob))
test.pred[test.pred.prob > 0.5] <- "high"

# confusion matrix
confusionMatrix(data = as.factor(test.pred), 
                reference = testing_data$mpg_cat, 
                positive = "high")
```

I decided on a decision rule of 0.5 to classify between "High" and "Low" efficiency cars. This is equivalent to assuming that there are equal consequences associated with assigning False Positives (misclassification of a "Low"-efficiency car as "High") and False Negatives (misclassification of a "High"-efficiency car as "Low").

With an overall accuracy of 0.897, MARS will correctly classify the "High"
 efficiency vehicles 85.7% of the time (Sensitivity) and correctly classifies the "Low" efficiency vehicles 93.4% of the time.